---
title: "COVID-19 Project; DATA 583"
author: "Yuening Li, Sneha Mikkilineni Durga, Binal Patel"
date: "3/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE)
library(dplyr)
library(forecast)
library(ggplot2)
```

# COVID-19 Time Series Analysis
#### Team members: Sneha Mikkilineni Durga, Binal Patel, Yuening Li


# Analysis of COVID-19 in Canada

##### Read in canada data

We split the data to training and testing; Training data: cases of COVID 19 in Canada From 2020-01-26 to 2020-04-18; Testing data: cases of COVID 19 in Canada From 2020-4-19 to 2020-4-27 

```{r}
## Training data: cases of COVID 19 in Canada From 2020-01-26 to 2020-04-18
canada <- read.csv("./Canada.csv")
canada$date = as.Date(canada$date, format = "%m/%d/%Y")
train<- canada[,c("date","total_cases")]

# Testing data: cases of COVID 19 in Canada From 2020-4-19 to 2020-4-27 
canada_predict <- read.csv("./canada_prediction.csv")
canada_predict$date = as.Date(canada_predict$ï..date, format = "%m/%d/%Y")
test <- canada_predict[,c("date","total_cases")]

```

##### inspect the training data

```{r,out.width= "50%", out.extra='style="float:right; padding:10px"'}
library(ggplot2)
ggplot(train, aes(y=total_cases, x=date)) +
  geom_point() +
  scale_x_date(date_labels = "%m/%d",date_breaks = "1 day")+
  xlab("date")+
  ylab("total cases")+
  ggtitle("Total cases of COVID 19 in Canada From 2020-01-26 to 2020-04-18") +
  theme_bw()+
  theme(axis.text.x = element_text(size = 5,angle = 90,hjust = 1))
```


##### Model 1. Arima (Auto regressive time series)

Making data frame into time series object

```{r}
# xts library used for time series 
library(xts)
case <- train$total_cases
names(case)<-as.character(train$date)
case<-as.xts(case)
```

test for stationary of time series

```{r,eval=FALSE}
library(tseries)
adf.test(case)
```

p-value = 0.99 suggest non-stationary, Next we use auto.arima() function to fit appropriate arima for the data, the function would make the time series stationary for us.

```{r}
library(forecast)
case_arima <- auto.arima(case)
case_arima
```

The appropriate model for canada covid 19 new cases is ARIMA(0,2,1) process, which means order 1 moving average and degree of defferencing 2.

Next, Use fitted time series model to
i. predict the next 8 days total cases of covid 19 , and
ii. compare the predicted values with the actual total cases from 2020-4-19 to 2020-4-27 by subtracting the actual values from the predictions and dividing by the estimated standard errors.

```{r}
nsteps = 9
case_pred <- predict(case_arima, n.ahead=nsteps)
(case_pred$pred - test$total_cases)/case_pred$se
```

All values are between -2 and 2, suggesting the model is quite accurate

```{r,out.width= "30%"}
test$pred<- case_pred$pred 
ggplot(test, aes(date)) + 
  geom_line(aes(y = total_cases, colour = "total_cases")) + 
  geom_line(aes(y = pred, colour = "pred"))+
  xlab("date")+
  ylab("total cases")+
  ggtitle("Actual and Predicted Total cases of COVID 19 in Canada")
```

The plot suggest we acquire quite accurate predictions.

##### Other Models

We convert the dates into "days since 2020/1/26" as this would give numeric values in form of 1, 2, 3, 4...

```{r}
# train data
x<-c(1:nrow(train))
y<- train$total_cases
new <- data.frame(x,y)
# test data
x<-c(1:nrow(test))
y<- test$total_cases
new_data <- data.frame(x,y)
```

###### Model 1. Quadratic Polynomial regression

```{r,out.width= "30%"}
library(DAAG)
case_lm2 <- lm(y ~ x + I(x^2), data = new)
plot(new, pch=16,xlab="Number of days from 2020-1-26",ylab = "Total cases",main = "Quadratic")
with(new, lines(x, predict(case_lm2), col=4, lwd=2))
```

The plot suggest we might need higher order polynomial model for better fit. 

###### Model 2. Higher order Polynomial Regression Model, WIth order = 4

```{r,out.width= "30%"}
# polynomial
case.poly.lm <- lm(y ~ poly(x,4),data=new)
plot(new, pch=16,xlab="Number of days from 2020-1-26",ylab = "Total cases", main = "General Polynomial with order 4")
with(new, lines(x, predict(case.poly.lm), col=3, lwd=2))
```

###### Model 3. Addictive (Penalized Spline) Model

Use the gam() function in the mgcv package to fit the additive Penalized Spline model

```{r,out.width= "30%"}
library(mgcv)
case_gam <- gam(y ~ s(x, k=6), data = new)
plot(new, pch=16,xlab="Number of days from 2020-1-26",ylab = "Total cases", main = "Penalized Spline")
with(new, lines(x, predict(case_gam), col=4, lwd=2))
```


###### Model 4. B-Spline

```{r,out.width= "30%"}
library(splines)
case_b_spl <- lm(y ~ bs(x,knots=c(40,50,55,60,65),degree=1,intercept=TRUE)-1,data=new)
plot(new, pch=16,xlab="Number of days from 2020-1-26",ylab = "Total cases", main = "B-Spline")
with(new, lines(x, predict(case_b_spl), col=4, lwd=2))
```

###### Model 5. GLM: Poisson regression

for count responses, like total number of covid 19 cases, A poisson or quasipoisson GLM is a reasonable guess for such data.

```{r,out.width= "30%"}
case_glm<- glm(y ~ x, family = quasipoisson, data = new)
plot(new, pch=16,xlab="Number of days from 2020-1-26",ylab = "Total cases", main = "GLM:Poisson")
with(new, lines(x, predict(case_glm,type="response"), col=4, lwd=2))
```

##### Model 6. Negative binomial Model

An alternative to quisilkelihood here is Negative binomial model, This is a model for counts which allows for overdispersion.

```{r}
library(MASS)
case_nb <- glm.nb(y~ x, data = new)
plot(new, pch=16,xlab="Number of days from 2020-1-26",ylab = "Total cases", main = "Negative binomial")
with(new, lines(x, predict(case_nb,type="response"), col=4, lwd=2))
```

The residual deviance is is close to the residual degrees of freedom, supporting the model

##### Which Model fits best?

Using the predict() function with the actual total cases from 2020-4-20 to 2020-4-27 as newdata object, compute the mean of the squares of the prediction errors for total cases to decide which of the above models is best for the data.(MSE)

```{r,out.width= "30%"}

# case_arima case.poly.lm  case_gam  case_b_spl  case_glm
# time series 22235232
print("time series MSE")
mean((predict(case_arima, n.ahead=9)$pred - new_data$y)^2)
# polynomial 13711510567
print("polynomial with order 4 MSE")
mean((predict(case.poly.lm, newdata=new_data) - new_data$y)^2)
# additive/Penalized Splines 13568357726
print("Penalized Splines Model MSE")
mean((predict(case_gam, newdata=new_data) - new_data$y)^2)
# B-spline 13605013369
print("B-spline Model MSE")
mean((predict(case_b_spl, newdata=new_data) - new_data$y)^2)
# GLM quasipoisson regresson 13605588889
print("GLM quasipoisson MSE")
mean((predict(case_glm, newdata=new_data) - new_data$y)^2)
print("GLM Negative binomial Model MSE")
mean((predict(case_glm, newdata=new_data) - new_data$y)^2)
```

We find that time series model perform much better than other models.


# Canada Provincial Data

##### Importing Data

```{r}
cases_prov <- read.csv("cases_timeseries_prov.csv")
cases_BC<- subset(cases_prov,cases_prov$province=='BC')
cases_BC<-cases_BC[,-c(1,3)]
cases_Alberta<- subset(cases_prov,cases_prov$province=='Alberta') 
cases_Alberta<-cases_Alberta[,-c(1,3)]
cases_Ont<- subset(cases_prov,cases_prov$province=='Ontario')
cases_Ont<-cases_Ont[,-c(1,3)]
cases_Quebec<- subset(cases_prov,cases_prov$province=='Quebec')
cases_Quebec<-cases_Quebec[,-c(1,3)]

```

##### Data exploration

```{r}
normalize <- function(x) {
return ((x - min(x)) / (max(x) - min(x)))
}
cases_prov <- read.csv("cases_timeseries_prov.csv")
cases_prov$date_report = as.Date(cases_prov$date_report, format = "%d-%m-%Y")
cases_prov%>%
    ggplot(aes(x=date_report, y=normalize(cases))) +  geom_line(aes(group=1))+
    facet_wrap( ~ province, scales = "free_x")+ylab("Cases")+xlab("Date")

```

##### Making data frame into time series object

```{r}
library(xts)
can_bcts=xts(cases_BC$cumulative_cases,as.Date(cases_BC$date_report,format='%d-%m-%Y'))
can_abts=xts(cases_Alberta$cumulative_cases,as.Date(cases_Alberta$date_report,format='%d-%m-%Y'))
can_onts=xts(cases_Ont$cumulative_cases,as.Date(cases_Ont$date_report,format='%d-%m-%Y'))
can_qts=xts(cases_Quebec$cumulative_cases,as.Date(cases_Quebec$date_report,format='%d-%m-%Y'))
```


##### Plotting Autocorrelation and Time Series plots to check for stationarity

```{r}
#BC
ggtsdisplay(can_bcts) #non stationary, decaying ACF

#Alberta
ggtsdisplay(can_abts) #non stationary, decaying ACF

#Ontario
ggtsdisplay(can_onts) #non stationary, decaying ACF

#Quebec
ggtsdisplay(can_qts) #non stationary, decaying ACF

```

Since the ACF plot shows a geometric decay and there is an increasing trend in the total number of cases, we know that the time series data is non stationary. Using a statistical test method, ADF we can clarify whether the time series is stationary or not. 

##### Applying Differencing to the four provinces

British Columbia 

```{r}
library(tseries)
difference<-na.omit(diff(can_bcts))
adf.test(can_bcts) #Non stationary pvalue > 0.05
adf.test(difference) #Non stationary pvalue >.05
diff2<-na.omit(diff(difference)) #Stationary after twice differencing pvalue < 0.05
adf.test(diff2)
```

Alberta 

```{r}
adf.test(can_abts) #Non stationary pvalue > 0.05
diff_ab<-na.omit(diff(can_bcts))
adf.test(diff_ab) #Non stationary pvalue >.05
diff2ab<-na.omit(diff(diff_ab)) #Stationary after twice differencing pvalue < 0.05
adf.test(diff2ab)
```

Ontario

```{r}
adf.test(can_onts) #Non stationary pvalue > 0.05
diff_on<-na.omit(diff(can_onts))
adf.test(diff_on) #Non stationary pvalue >.05
diff2on<-na.omit(diff(diff_on)) #Stationary after twice differencing pvalue < 0.05
adf.test(diff2on)
```

Quebec

```{r}
adf.test(can_qts) #Non stationary pvalue > 0.05
diff_q<-na.omit(diff(can_qts))
adf.test(diff_q) #Non stationary pvalue >.05
diff2q<-na.omit(diff(diff_q)) #Stationary after twice differencing pvalue < 0.05
adf.test(diff2q)
```

In order to make the time series stationary, we had to take a difference twice to get a p-value less than 0.05. When we use the ARIMA model, we will be using d=2 as our parameter.

##### ARIMA Model for all the four provinces

Now using Arima model to forecast number of cases in the next 20 days

British Columbia

```{r}
library(ggplot2)
dd<-auto.arima(can_bcts)
dd
forcastfit.bc <- Arima(can_bcts, order=c(1,2, 2))
summary(forcastfit.bc)
f=forecast(forcastfit.bc, n.ahead=20)
plot(forecast(forcastfit.bc,h=20),xlab = "Days",ylab="Number of Cumulative Cases")
forecast(forcastfit.bc,20)
#ggtsdiag(forcastfit.ar) #plotting residuals

```


Alberta

```{r}

dd<-auto.arima(can_abts)
dd
forcastfit.ab <- Arima(can_abts, order=c(1, 2, 0))
summary(forcastfit.ab)
f=forecast(forcastfit.ab, n.ahead=20)
plot(forecast(forcastfit.ab,h=20),xlab = "Days",ylab="Number of Cumulative Cases")
forecast(forcastfit.ab,20)
#ggtsdiag(forcastfit.ar) #plotting residuals

```


Ontario

```{r}
dd<-auto.arima(can_onts)
dd
forcastfit.on <- Arima(can_onts, order=c(1, 2, 0))
summary(forcastfit.on)
f=forecast(forcastfit.on, n.ahead=20)
plot(forecast(forcastfit.on,h=20),xlab = "Days",ylab="Number of Cumulative Cases")
forecast(forcastfit.on,20)

```


Quebec

```{r}
dd<-auto.arima(can_qts)
dd
forcastfit.q <- Arima(can_qts, order=c(1, 2, 3))
summary(forcastfit.q)
f=forecast(forcastfit.q, n.ahead=20)
plot(forecast(forcastfit.q,h=20),xlab = "Days",ylab="Number of Cumulative Cases")
forecast(forcastfit.q,20)

```




